\myTitle

### 2015-03-12 tex/response_to_refs_letter.tex

Thank you for the opportunity to resubmit our revised manuscript, MER-14-0343 ``Interpreting the FLOCK algorithm
from a statistical perspective.''  We are very grateful to you and to both of the referees for the many
suggestions that we have incorporated.  We feel they have improved the manuscript greatly.  We have taken your
advice and undertaken a set of simulations that provide a thorough assessment of {\sc flock} versus {\sc 
structure} on data sets with large numbers (2,000) of individuals, like our original steelhead data set.
We have dropped the steelhead data set from the paper.

In order to conduct these simulations we implemented the {\sc flock} algorithm by modifying the source code of
the program {\sc structure}.  We used the resulting program, {\sc flockture}, because {\sc flock} itself
is 200 times slower than {\sc flockture} and it is also not amenable to scripting and repeated application to 
multiple data sets in batch mode.  It is simply not possible to prepare a reproducible set of simulations using
the point-and-click interface of {\sc flock}.  We show simulations in the paper to verify that {\sc flockture} and {\sc flock} yield equivalent results before comparing {\sc structure} to {\sc flockture}.

Another change in the paper is that we focus on the performance of the programs for clustering individuals while
assuming the correct value of $K$.  We were actually somewhat uncomfortable in our initial submission that
assessing {\sc flock}'s ability to estimate $K$ might give the impression that we recommend such pursuits---an
endeavor that I (along with {\sc structure} co-author Matthew Stephens) are reluctant to recommend, especially
when population structure is weak.  So, we are pleased to be able to offer a rigorous evaluation of program
performance without going too deeply into the {\em ad-hoc} procedures for estimating $K$ with {\sc flock}.

Overall, we discovered through our simulations that results from the {\sc flock} algorithm tend to be less 
consistent than {\sc structure}'s, especially at low values of $F_\mathrm{ST}$.  But, on the other hand, we were 
quite interested and pleased to see that the algorithm still works quite well in some cases.  In fact, it outperformed {\sc structure} 
using SNPs at intermediate levels of differention.  We also found that in large data sets like those that
we simulated, {\sc flock}'s individual runs seldom (or never) converge to exactly the same partition.  This 
presents some serious problems for applying the Duchesne \& Turgeon (2012) recipe for estimating $K$, and we 
note that in the discussion.

@@

Below, we address individual comments from you and the referees.  Our replies are in \textcolor{blue}{{\sl blue slant}} font.

Thank you again for your service.  As usual, we have been greatly pleased with the review process
at {\em Molecular Ecology Resources} and are excited with how much improvement the referees have brought
to our paper.  We apologize that it has required several months to return these revisions, but do appreciate
the extra time the journal granted us---the extra time was necessary to write {\sc flockture} and to undertake the
simulations.

@@

They should be published in a full-fledged article, not just a comment.

\reply{If, upon reading the present version of our manuscript, the editor feels it would be appropriate
to publish it as a full article, we would agree to that.
}

I also suggest that the authors make an additional effort to better understand plateau analysis especially in connection with variability of solutions i.e. lengths of plateaus. Short plateaus (numerous solutions) are the right answer whenever K is the wrong answer to the `number of populations' problem.

\reply{We have taken the referee's suggestion and worked on obtaining a better understanding of plateau analysis.
Short plateaus may well be right answer whenever $K$ is the wrong answer.  However, short plateaus are also
the wrong answer when the $K$ value is correct.  This seems to be a major problem for {\sc flock} on data sets
of the size that we have simulated.  We've addessed this in the 2nd to last paragraph of the discussion. }

### 2015-03-12 tex/abstract.tex
Addressing reviewer comments.
Revised version as submitted.

      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
interpreted as an estimation procedure based on 
a model essentially identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. Rather than using MCMC, the {\sc flock} algorithm 
searches for the maximum-{\em a-posteriori}
estimate of this {\sc structure} model via a simulated 
annealing algorithm with a rapid cooling 
schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  We 
demonstrate the similarities between the two programs in a two step approach. First,
to enable rapid batch processing of many simulated data sets,
we modified the source code of  {\sc structure} to use the {\sc flock} algorithm, producing
the program {\sc flockture}. With simulated data we confirmed that results obtained with
{\sc flock}  and {\sc flockture} and very similar (though flockture is some 200 times faster). 
Second, we simulated multiple large datasets under varying 
levels of population differentiation for both microsatellite and SNP genotypes. We analyzed them
with {\sc flockture} and {\sc structure} and assessed each program on its ability to correctly cluster
individuals to their correct subpopulation.  We show that
{\sc flockture} yields results similar to {\sc structure} although with greater 
variability from run to run. {\sc flockture} did perform better than {\sc structure} 
when genotypes were composed of SNPs and differentiation was moderate 
($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} outperformed {\sc flockture}
for both marker types. Interpreting {\sc flock}'s algorithm as a special case of the model in 
{\sc structure} should aid in understanding the program's output and behavior.

>>>>

      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
interpreted as an estimation procedure based on 
a model essentially identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. Rather than using MCMC, the {\sc flock} algorithm 
searches for the maximum-{\em a-posteriori}
estimate of this {\sc structure} model via a simulated 
annealing algorithm with a rapid cooling 
schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  We 
demonstrate the similarities between the two programs in a two step approach. First,
to enable rapid batch processing of many simulated data sets,
we modified the source code of  {\sc structure} to use the {\sc flock} algorithm, producing
the program {\sc flockture}. With simulated data we confirmed that results obtained with
{\sc flock}  and {\sc flockture} and very similar (though flockture is some 200 times faster). 
Second, we simulated multiple large datasets under varying 
levels of population differentiation for both microsatellite and SNP genotypes. We analyzed them
with {\sc flockture} and {\sc structure} and assessed each program on its ability to cluster
individuals to their correct subpopulation.  We show that
{\sc flockture} yields results similar to {\sc structure} albeit with greater 
variability from run to run. {\sc flockture} did perform better than {\sc structure} 
when genotypes were composed of SNPs and differentiation was moderate 
($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} outperformed {\sc flockture}
for both marker types. On large data sets like those we simulated, it appears that 
{\sc flock}'s reliance on inference rules regarding its ``plateau record'' are not helpful. 
Interpreting {\sc flock}'s algorithm as a special case of the model in 
{\sc structure} should aid in understanding the program's output and behavior.

### 2015-03-12 tex/main-body-text.tex
Revised version as submitted.

To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
were simulated using a coalescent-based program (code provided at 
https://github.com/eriqande/flock-comment/). The 5 populations
were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
Each of the four data sets was simulated with a different 
average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
Both {\sc flock} and {\sc flockture} were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into correct population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

@@

Plateaus were only observed for the dataset 
with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
the largest plateau (Supp. Table 1). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
variability that we would expect from randomizing the initial grouping of individuals to 
initiate the algorithms.

@@

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{
  Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flock}). 
  Individuals are grouped left to right into their simulated population samples, and each 
  horizontal plot represents a run of the program
  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flock} (unless
  there were no plateaus, in which case the run reaching the highest likelihood was chosen).
  Within horizontal plots, each individual is represented by a vertical bar composed of colored
  segments of lengths proportional to the various cluster membership probabilities.
  Each horizontal plot
  represents one of nine 
 runs of the programs  at $K$ = 5 (the true value of $K$). Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Numbers in 
parentheses following the model names give the number, out of 9, of replicate runs
that obtained the same visually-indistinguishable estimate. (Note that although many of {\sc flocktures}'s solutions in B--D were visually indistinguishable, they were not identical partitions, and would
hence be considered different for the purposes of creating a {\sc flock} plateau record.) Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s solutions consistently assigned fewer individuals correctly---and the
replicate solutions for individual data sets became more variable---at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
be multimodal---possessing numerous local peaks and troughs.
The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} and {\sc flock} to settle on settle on different solutions in 
different runs. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for the {\sc flock} algorithm.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

@@

The above presents a serious conundrum to the user of {\sc flock}.  On the one hand,
{\sc flock} clearly continues to provide useful inference about population structure
(as evidenced by the {\sc distruct} plots in Fig.~\ref{fig:FvSdistruct}), even when its
plateau record includes no plateaus; however, on the other hand, if there were no
plateaus for $K=5$ in our simulated data, then the {\sc flock} method would lead to an estimate of $K$ that
was either "undecided" or some value less than $K$. The authors of {\sc flock}
discourage users from making any inference at a value of $K$ that is not
the value estimated by {\sc flock}, because, ``outside the estimated $K$, no partitions from {\sc flock}
 should be considered. Therefore\ldots clusterings from {\sc flock}\ldots for all $K$ except the estimated one are just
 irrelevant and not valid''(Pierre Duchesne, pers. comm).  To be specific, following 
this advice of {\sc flock}'s authors would have led to discarding {\sc flockture}'s
results for any $F_\mathrm{ST} < 0.032$ (almost 90\% of our simulated data sets!), even those that are
manifestly quite accurate---some competitive
with or even superior to {\sc structure}'s estimate!  We surmise that this problem with {\sc flock} 
guidance issues from the fact that the {\em ad-hoc} rules developed by
\citeauthor{Duc&Tur2012} for estimating $K$ with {\sc flock} were developed empirically
by reference to a small set of simulations over a fairly limited range of sample sizes.  Since our simulated
samples (made to mimic the sorts of large, real data sets we collect in our labs)
fall outside the range of values explored
by \citeauthor{Duc&Tur2012} it appears that {\sc flock}'s rules of use are inappropriate, 
and perhaps over-restrictive for these data sets.  Researchers with data that similarly fall beyond the range of 
data sets tested in \citeauthor{Duc&Tur2012} may wish to approach {\sc flock} analysis with
caution, or may choose to rely on other programs.

>>>>

To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
were simulated using the coalescent program {\sc makesamples} \citep{Hudson2002} (code provided at 
https://github.com/eriqande/flock-comment/). The 5 populations
were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
Each of the four data sets was simulated with a different 
average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
Both {\sc flock} and {\sc flockture} were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into correct population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

@@

Plateaus were only observed for the dataset 
with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
the largest plateau. Multiple runs of {\sc flockture} demonstrated that the differences are well within the
variability that we would expect from randomizing the initial grouping of individuals to 
initiate the algorithms.

@@

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{
  Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flockture}). 
  Individuals are grouped left to right into their simulated population samples, and each 
  horizontal plot represents a run of the program
  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flockture} (unless
  there were no plateaus, in which case the run reaching the highest likelihood was chosen).
  Within horizontal plots, each individual is represented by a vertical bar composed of colored
  segments of lengths proportional to the various cluster membership probabilities.
  Each horizontal plot
  represents one of nine 
 runs of the programs  at $K$ = 5 (the true value of $K$). Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Numbers in 
parentheses following the model names give the number, out of 9, of replicate runs
that obtained the same visually-indistinguishable estimate. (Note that although many of {\sc flocktures}'s solutions in B--D were visually indistinguishable, they were not identical partitions, and would
hence be considered different for the purposes of creating a {\sc flock} plateau record.) In E, all 9
of {\sc flocktures}'s solutions were visually distinguishable, though we show only one.  Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s solutions consistently assigned fewer individuals correctly---and the
replicate solutions for individual data sets became more variable---at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
multimodal---possessing numerous local peaks and troughs.
The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} and {\sc flock} to settle on settle on different solutions in 
different runs. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for the {\sc flock} algorithm.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

@@

The above presents a serious conundrum to the user of {\sc flock}.  On the one hand,
{\sc flock} clearly continues to provide useful inference about population structure
(as evidenced by the {\sc distruct} plots in Fig.~\ref{fig:FvSdistruct}), even when its
plateau record includes no plateaus; however, on the other hand, if there were no
plateaus for $K=5$ in our simulated data, then the {\sc flock} method would lead to an estimate of $K$ that
was either "undecided" or some value less than $K$. The authors of {\sc flock}
discourage users from making any inference at a value of $K$ that is not
the value estimated by {\sc flock}, because, ``outside the estimated $K$, no partitions from {\sc flock}
 should be considered. Therefore\ldots clusterings from {\sc flock}\ldots for all $K$ except the estimated one are just
 irrelevant and not valid''(Pierre Duchesne, pers. comm).  To be specific, following 
this advice of {\sc flock}'s authors would have led to discarding {\sc flockture}'s
results for any $F_\mathrm{ST} < 0.032$ (almost 90\% of our simulated data sets); even those that are
manifestly quite accurate---some competitive
with or even superior to {\sc structure}'s estimate.  We surmise that this problem with the guidance on 
{\sc flock} stems from the fact that the {\em ad-hoc} rules developed by
\citeauthor{Duc&Tur2012} for estimating $K$ with {\sc flock} were developed empirically
by reference to a small set of simulations over a fairly limited range of sample sizes.  Since our simulated
samples (made to mimic the sorts of large, real data sets we collect in our labs)
fall outside the range of values explored
by \citeauthor{Duc&Tur2012}, it appears that {\sc flock}'s rules of use are inappropriate, 
and perhaps over-restrictive for these data sets.  Researchers with data that similarly fall beyond the range of 
data sets tested in \citet{Duc&Tur2012} may wish to approach {\sc flock} analysis with
caution, or may choose to rely on other programs.
